---
title: "Fbhighres"
author: "A.DeMartin"
date: "2025-04-16"
output: 
  html_document:
    keep_md: true
    toc: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
options(width = 100)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dev = c("png", "pdf"))
seed <- 1234
```

## load packages
```{r load packages, warning=FALSE, include=FALSE}
library(ExploreSCdataSeurat3)
library(runSeurat3)
library(Seurat)
library(ggpubr)
library(pheatmap)
library(SingleCellExperiment)
library(dplyr)
library(tidyverse)
library(viridis)
library(muscat)
library(circlize)
library(destiny)
library(scater)
library(metap)
library(multtest)
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)
library(enrichplot)
library(DOSE)
library(grid)
library(gridExtra)
library(ggupset)
library(VennDiagram)
library(NCmisc)
```

## load object
```{r}
fileNam <- "/Users/immbio/Desktop/Project/Angelina/HumanHeartCarTrans2/data/Human_heart_allmerged_seurat.rds"
seuratM <- readRDS(fileNam)
```

## set color vectors 
```{r}
colclusterName <- c("#67001f", "#f4a582","#D53E4F", "#FEE08B","#8c510a","#003c30","#01665e","#66C2A5", "#3288BD","#BEAED4", "#c7eae5", "#B09C85", "#4e5a4c","#393A3F","#355C7D","#202547","#B45B5C","#725663FF","#232429","#bac1d1")
names(colclusterName) <- c("CM","Fb1","Fb2","Fb3","PerivFb1","PerivFb2","VSMC","BEC1","BEC2","BEC3","LEC","NC","Baror","Adipoc","Mph1","Mph2","Tcell1","Tcell2","Mastcell","Mph3")

coldiseaseCond <- c("#dfc27d","#BE3144","#202547","#355C7D","#779d8d")
names(coldiseaseCond) <- c("healthy", "explant", "visit1", "visit2", "visit3")
```

```{r}
## check umap
Idents(seuratM) <- seuratM$clusterName
DimPlot(seuratM, reduction = "umap", cols = colclusterName, label= TRUE)
```

## subset Fb, PerivFb, VSMC (AllFb)
```{r}
## subset Fb, PerivFb and VSMC
seuratAllFb <- subset(seuratM, idents = c("Fb1", "Fb2", "PerivFb1", "VSMC")) ## exclude Fb3 and PerivFb2 - int cells/doublets
levels(seuratAllFb)
table(seuratAllFb$orig.ident)
table(seuratAllFb$clusterName)
DimPlot(seuratAllFb, reduction = "umap", cols = colclusterName, label=TRUE)
```

## rerun seurat AllFb 
```{r, eval=FALSE, include=TRUE}
## rerun seurat
seuratAllFb <- NormalizeData (object = seuratAllFb)
seuratAllFb <- FindVariableFeatures(object = seuratAllFb)
seuratAllFb <- ScaleData(object = seuratAllFb, verbose = TRUE)
seuratAllFb <- RunPCA(object=seuratAllFb, npcs = 30, verbose = FALSE)
seuratAllFb <- RunTSNE(object=seuratAllFb, reduction="pca", dims = 1:20)
seuratAllFb <- RunUMAP(object=seuratAllFb, reduction="pca", dims = 1:20)
seuratAllFb <- FindNeighbors(object = seuratAllFb, reduction = "pca", dims= 1:20)

res <- c(0.25, 0.6, 0.8, 0.4)
for (i in 1:length(res)) {
  seuratAllFb <- FindClusters(object = seuratAllFb, resolution = res[i], random.seed = 1234)
}

table(seuratAllFb$orig.ident)
table(seuratAllFb$RNA_snn_res.0.25)
```

## load file AllFb
```{r}
fileNam <- "/Users/immbio/Desktop/Project/Angelina/HumanHeartCarTrans2/data/Human_heart_AllFb.rds"
seuratAllFb <- readRDS(fileNam)
table(seuratAllFb$dataset)
table(seuratAllFb$RNA_snn_res.0.25)
table(seuratAllFb$orig.ident)
```

```{r, include=TRUE, eval=FALSE}
#### cluster_name_Fb
seuratAllFb$clusterNameAllFb <- "clusterNameAllFb"
seuratAllFb$clusterNameAllFb[which(seuratAllFb$RNA_snn_res.0.25 %in% "0" )] <- "Fb1"
seuratAllFb$clusterNameAllFb[which(seuratAllFb$RNA_snn_res.0.25 %in% "1" )] <- "Fb2"
seuratAllFb$clusterNameAllFb[which(seuratAllFb$RNA_snn_res.0.25 %in% "2" )] <- "Fb3"
seuratAllFb$clusterNameAllFb[which(seuratAllFb$RNA_snn_res.0.25 %in% "3" )] <- "PerivFb1"
seuratAllFb$clusterNameAllFb[which(seuratAllFb$RNA_snn_res.0.25 %in% "4" )] <- "PerivFb2"
seuratAllFb$clusterNameAllFb[which(seuratAllFb$RNA_snn_res.0.25 %in% "5" )] <- "Fb4"
seuratAllFb$clusterNameAllFb[which(seuratAllFb$RNA_snn_res.0.25 %in% "6" )] <- "VSMC"
seuratAllFb$clusterNameAllFb[which(seuratAllFb$RNA_snn_res.0.25 %in% "7" )] <- "Fb5"

table(seuratAllFb$clusterNameAllFb)
table(seuratAllFb$RNA_snn_res.0.25)

###order
Idents(seuratAllFb) <- seuratAllFb$clusterNameAllFb
seuratAllFb$clusterNameAllFb <- factor(seuratAllFb$clusterNameAllFb, levels=c("Fb1", "Fb2", "Fb3", "Fb4", "Fb5", "PerivFb1", "PerivFb2", "VSMC"))
Idents(seuratAllFb) <- seuratAllFb$clusterNameAllFb
table(seuratAllFb$clusterNameAllFb)
```

```{r, eval=FALSE, include=TRUE}
### save seurat object
saveRDS(seuratAllFb, file="/Users/immbio/Desktop/Project/Angelina/HumanHeartCarTrans2/data/Human_heart_AllFb.rds")
```

##set color vectors AllFb
```{r set color vector AllFb}
colAllFb <- c("#D53E4F","#f4a582","#ff7b7b","#8e0b00","#FEE08B","#42090D","#FF7B00","#FFF4DF")
names(colAllFb) <- c("Fb1","PerivFb1","Fb2","Fb3","VSMC","Fb4","PerivFb2", "Fb5")
```

## umaps
```{r umaps AllFb}
Idents(seuratAllFb) <- seuratAllFb$RNA_snn_res.0.25
DimPlot(seuratAllFb, reduction = "umap", pt.size = 0.5, label = TRUE)

Idents(seuratAllFb) <- seuratAllFb$clusterName
DimPlot(seuratAllFb, reduction = "umap", pt.size = 0.5, cols = colclusterName, label = TRUE)

Idents(seuratAllFb) <- seuratAllFb$clusterNameAllFb
DimPlot(seuratAllFb, reduction = "umap", pt.size = 0.5, cols = colAllFb, label = TRUE)
DimPlot(seuratAllFb, reduction = "umap", pt.size = 0.5, cols = colAllFb)  + theme(legend.position = "null")

Idents(seuratAllFb) <- seuratAllFb$diseaseCond
DimPlot(seuratAllFb, reduction = "umap", pt.size = 0.5, cols = coldiseaseCond, shuffle= TRUE)
DimPlot(seuratAllFb, reduction = "umap", pt.size = 0.5, cols = coldiseaseCond, shuffle = TRUE) + theme(legend.position = "null")
```
## calculate cluster marker genes AllFb
```{r, include=TRUE, eval=FALSE}
## cluster marker
Idents(seuratAllFb) <- seuratAllFb$clusterNameAllFb
markerGenes <- FindAllMarkers(seuratAllFb, only.pos=T) %>% 
  dplyr::filter(p_val_adj < 0.01)

## save table
write.table(markerGenes, 
            file= "/Users/immbio/Desktop/Project/Angelina/HumanHeartCarTrans2/data/markerGenesAllFb_clusterNameAllFb",
            sep="\t",
            quote=F,
            row.names=F,
            col.names=T)
```

### BMP features 
```{r features BMP AllFb}
genes <- data.frame(gene=rownames(seuratAllFb)) %>% 
  mutate(geneID=gsub("^.*\\.", "", gene))

selGenes <- data.frame(geneID=c("HGF", "MET", "GREM1", "GREM2", "BMPR1A", "BMPR2", "BMP2K", "BMP8A", "BMP1", "BMP6", "BMP2", "BMP5", "BMP4")) %>% 
  left_join(., genes, by="geneID")

pList <- sapply(selGenes$gene, function(x){
  p <- FeaturePlot(seuratAllFb, features = x, 
                   reduction = "umap", pt.size = 0.1, 
                   cols = c("lightgrey", "#BE3144"), 
                   raster = FALSE) +
    theme(legend.position="right")
  plot(p)
})
```
### BMP featuresb ordered
```{r features BMP ordered AllFb}
genes <- data.frame(gene=rownames(seuratAllFb)) %>% 
  mutate(geneID=gsub("^.*\\.", "", gene))

selGenes <- data.frame(geneID=c("HGF", "MET", "GREM1", "GREM2", "BMPR1A", "BMPR2", "BMP2K", "BMP8A", "BMP1", "BMP6", "BMP2", "BMP5", "BMP4")) %>% 
  left_join(., genes, by="geneID")

pList <- sapply(selGenes$gene, function(x){
  p <- FeaturePlot(seuratAllFb, features = x, 
                   reduction = "umap", pt.size = 0.1, 
                   cols = c("lightgrey", "#BE3144"), 
                   raster = FALSE, 
                   order=TRUE) +
    theme(legend.position="right")
  plot(p)
})
```

### BMP violin plots
```{r violin plots BMP AllFb}
Idents(seuratAllFb) <- seuratAllFb$diseaseCond
pList <- sapply(selGenes$gene, function(x){
  p <- VlnPlot( object = seuratAllFb, features = x, pt.size = 1, cols=coldiseaseCond)
  plot(p)
})
```

### violin plots
```{r violin plots AllFb}
Idents(seuratAllFb) <- seuratAllFb$diseaseCond
pList <- sapply(selGenes$gene, function(x){
  p <- VlnPlot( object = seuratAllFb, features = x, pt.size = 0, cols=coldiseaseCond)
  plot(p)
})
```

## abundance plots
### dataset
```{r abundance dataset AllFb, fig.height=10, fig.width=10}
order_keywords <- c("HH", "RV|LV|expLV|expRV|331571_3-5_20231012_Hu_nucseq_USZ_HTx001|331571_4-6_20231012_Hu_nucseq_USZ_HTx002", "V1", "V2","V3")
files <- unique(seuratAllFb$dataset)
ordered_files <- c()
for (key in order_keywords) {
  ordered_files <- c(ordered_files, files[grepl(key, files)])
}

###dataset
datList <- NULL
for(con in unique(seuratAllFb$dataset)){
  seuratSub <- subset(seuratAllFb, dataset==con)
  print(dim(seuratSub))
  dat_con <- as.data.frame(table(seuratSub$clusterNameAllFb)) %>%
  mutate(percent=Freq/ncol(seuratSub)) %>% mutate(dataset=con)
  datList[[con]] <- dat_con
}
dat_all <- do.call("rbind", datList)

## plot abundance
ggbarplot(dat_all, x= "dataset", y= "percent", fill = "Var1", palette = colAllFb, legend = "right", legend.titel = "cluster", ylab = "frequency")  + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + scale_x_discrete(limits=ordered_files)
```

### patient_diseaseCond
```{r abundance patient_diseaseCond AllFb, fig.height=8, fig.width=10}
order_keywords <- c("healthy", "visit1", "visit2", "visit3", "explant")
files <- unique(seuratAllFb$patient_diseaseCond)
ordered_files <- c()
for (key in order_keywords) {
  ordered_files <- c(ordered_files, files[grepl(key, files)])
}

###patient_diseaseCond
datList <- NULL
for(con in unique(seuratAllFb$patient_diseaseCond)){
  seuratSub <- subset(seuratAllFb, patient_diseaseCond==con)
  print(dim(seuratSub))
  dat_con <- as.data.frame(table(seuratSub$clusterNameAllFb)) %>%
  mutate(percent=Freq/ncol(seuratSub)) %>% mutate(patient_diseaseCond=con)
  datList[[con]] <- dat_con
}
dat_all <- do.call("rbind", datList)

## plot abundance
ggbarplot(dat_all, x= "patient_diseaseCond", y= "percent", fill = "Var1", palette = colAllFb, legend = "right", legend.titel = "cluster", ylab = "frequency")  + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + scale_x_discrete(limits=ordered_files)
```

### diseaseCond
```{r abundance diseaseCoond AllFb, fig.height=8, fig.width=10}
orddiseaseCond <- c("healthy","visit1", "visit2" ,"visit3", "explant")

###diseaseCond
datList <- NULL
for(con in unique(seuratAllFb$diseaseCond)){
  seuratSub <- subset(seuratAllFb, diseaseCond==con)
  print(dim(seuratSub))
  dat_con <- as.data.frame(table(seuratSub$clusterNameAllFb)) %>%
  mutate(percent=Freq/ncol(seuratSub)) %>% mutate(diseaseCond=con)
  datList[[con]] <- dat_con
}
dat_all <- do.call("rbind", datList)

## plot abundance
ggbarplot(dat_all, x= "diseaseCond", y= "percent", fill = "Var1", palette = colAllFb, legend = "right", legend.titel = "cluster", ylab = "frequency")  + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + scale_x_discrete(limits=orddiseaseCond)
```

## subset IntFb
```{r}
## subset IntFb
Idents(seuratAllFb) <- seuratAllFb$clusterNameAllFb
seuratIntFb <- subset(seuratAllFb, idents = c("Fb1", "Fb2", "Fb3", "Fb4", "Fb5"))
table(seuratIntFb$clusterNameAllFb)

DimPlot(seuratIntFb, reduction = "umap", cols = colAllFb, label=TRUE)
```

## rerun seurat IntFb
```{r, eval=FALSE, include=TRUE}
## rerun seurat
seuratIntFb <- NormalizeData (object = seuratIntFb)
seuratIntFb <- FindVariableFeatures(object = seuratIntFb)
seuratIntFb <- ScaleData(object = seuratIntFb, verbose = TRUE)
seuratIntFb <- RunPCA(object=seuratIntFb, npcs = 30, verbose = FALSE)
seuratIntFb <- RunTSNE(object=seuratIntFb, reduction="pca", dims = 1:20)
seuratIntFb <- RunUMAP(object=seuratIntFb, reduction="pca", dims = 1:20)
seuratIntFb <- FindNeighbors(object = seuratIntFb, reduction = "pca", dims= 1:20)

res <- c(0.25, 0.6, 0.8, 0.4, 0.1)
for (i in 1:length(res)) {
  seuratIntFb <- FindClusters(object = seuratIntFb, resolution = res[i], random.seed = 1234)
}

table(seuratIntFb$orig.ident)
table(seuratIntFb$RNA_snn_res.0.25)
```

```{r, include=TRUE, eval=FALSE}
## clustername IntFb
seuratIntFb$clusterNameIntFb <- "clusterNameIntFb"
seuratIntFb$clusterNameIntFb[which(seuratIntFb$RNA_snn_res.0.1 %in% "0" )] <- "Fb1/2"
seuratIntFb$clusterNameIntFb[which(seuratIntFb$RNA_snn_res.0.1 %in% "1" )] <- "Fb3"
seuratIntFb$clusterNameIntFb[which(seuratIntFb$RNA_snn_res.0.1 %in% "2" )] <- "Fb4"
seuratIntFb$clusterNameIntFb[which(seuratIntFb$RNA_snn_res.0.1 %in% "3" )] <- "Fb5"

table(seuratIntFb$clusterNameIntFb)
table(seuratIntFb$RNA_snn_res.0.1)

## order
Idents(seuratIntFb) <- seuratIntFb$clusterNameIntFb
seuratIntFb$clusterNameIntFb <- factor(seuratIntFb$clusterNameIntFb, levels=c("Fb1/2", "Fb3", "Fb4", "Fb5"))
Idents(seuratIntFb) <- seuratIntFb$clusterNameIntFb
table(seuratIntFb$clusterNameIntFb)
```

```{r, eval=FALSE, include=TRUE}
### save seurat object
saveRDS(seuratIntFb, file="/Users/immbio/Desktop/Project/Angelina/HumanHeartCarTrans2/data/Human_heart_IntFb.rds")
```

## load file IntFb
```{r}
fileNam <- "/Users/immbio/Desktop/Project/Angelina/HumanHeartCarTrans2/data/Human_heart_IntFb.rds"
seuratIntFb <- readRDS(fileNam)
table(seuratIntFb$clusterNameIntFb, seuratIntFb$diseaseCond)
```

```{r set color vector IntFb}
colIntFb <- c("#D53E4F","#8e0b00","#42090D","#FFF4DF")
names(colIntFb) <- c("Fb1/2","Fb3","Fb4", "Fb5")
```

## umaps
```{r umaps IntFb}
Idents(seuratIntFb) <- seuratIntFb$RNA_snn_res.0.1
DimPlot(seuratIntFb, reduction = "umap", pt.size = 0.5, label = TRUE)

Idents(seuratIntFb) <- seuratIntFb$clusterNameAllFb
DimPlot(seuratIntFb, reduction = "umap", pt.size = 0.5, cols = colAllFb, label = TRUE, shuffle = TRUE)

Idents(seuratIntFb) <- seuratIntFb$clusterNameIntFb
DimPlot(seuratIntFb, reduction = "umap", pt.size = 0.5, cols = colIntFb, label = TRUE, shuffle = TRUE)

Idents(seuratIntFb) <- seuratIntFb$diseaseCond
DimPlot(seuratIntFb, reduction = "umap", pt.size = 0.5, cols = coldiseaseCond, shuffle = TRUE)
```

### diffusion map
```{r}
pca_embeddings <- Embeddings(seuratIntFb, reduction = "pca")
dm <- DiffusionMap(pca_embeddings)

df <- data.frame(DC1 = eigenvectors(dm)[,1], DC2 = eigenvectors(dm)[,2])

## color according to clusterAllFb
df$cluster <- seuratIntFb$clusterNameAllFb

ggplot(df, aes(x = DC1, y = DC2, color = cluster)) +
  geom_point(size = 0.5) +
  scale_color_manual(values = colAllFb)

## color according to diseaseCond
df$cluster <- seuratIntFb$diseaseCond

ggplot(df, aes(x = DC1, y = DC2, color = cluster)) +
  geom_point(size = 0.5) +
  scale_color_manual(values = coldiseaseCond)
```

## session info
```{r date and session info}
date()
sessionInfo()
```
